Overrelaxed Sinkhorn--Knopp Algorithm for Regularized Optimal Transport

### 2017-10-20 paper.tex
Relecture Nicolas

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=7cm]{images/cvgce_zone}
\caption{\label{phi_omega} Value of $\varphi_\omega(x)$. The function is positive above the black line, negative below. For any relaxation parameter $\omega$ smaller than $2$, there exists a neighborhood of $1$ on which $\varphi_\omega(\cdot)$ is positive.}
%\includegraphics[width=11cm]{phi_omega.png}
%\caption{\label{phi_omega} Plot of functions $\varphi_\omega(x)$ for different values $1\leq \omega\leq 2$ and   $x\in[10^{-2}, 10^1]$.}
\end{center}
\end{figure}

@@

\begin{proposition}\label{prop:thetachoice}
	The function
	\begin{equation}
	\label{eq:theta_k}
	\theta_k(\gamma) =\Theta\left ((A_k \gamma)\oslash \mu^k\right)
	\end{equation}
	is continuous and verifies the descent condition (\ref{eq:cond_theta_k}).
\end{proposition}
This construction has several advantages:
\begin{itemize}
\item it allows to choose arbitrarily the parameter $\theta_0$ that will be used eventually when the algorithm is close to convergence (we motivate what are good choices for $\theta_0$ in section \ref{section:local});
\item it is also an easy approach to having an adaptive method, as the approximation of $\Theta^*$ has a negligible cost (it only require to solve a one dimensional problem that depends on the smallest value of $(A_k \gamma)\oslash \mu^k$, which can be done in a few iterations of Newton's method).
\item experiments show that this is often an excellent choice in practice.
\end{itemize}

@@

The resulting algorithm, which is proved to be convergent by Theorem~\ref{thm:algo}, is written in pseudo-code in Algorithm \ref{SOR}. It uses the function $\Theta$ defined implicitly in \eqref{eq:Theta}, which in practice is approximated with a few iterations of Newton's method on the function $\omega \mapsto \varphi_\omega(\min u)$ which decreasing is as can be seen on Figure~\ref{phi_omega}. With the choice $\theta_0=1$, one recovers exactly the original SK algorithm.

\begin{algorithm}
\caption{Overrelaxed SK algorithm }
\label{SOR}
\begin{algorithmic}
\REQUIRE $\mu^1\in \IR^{n_1}$, $\mu^2\in \IR^{n_2}$, $c\in \IR^{n_1\times n_2}_+$
\STATE Set $a=\mathbf{1}_{n_1}$, $b=\mathbf{1}_{n_2}$, $\gamma^0=e^{-c/\epsilon}$, $\theta_0\in[1;2)$ and $\eta>0$
\WHILE {$||a\otimes \gamma^0 b -  \mu_1||>\eta$}
\STATE $\tilde a=\mu_1\oslash (\gamma^0 b)$, 
\STATE $\omega=\Theta(a\oslash\tilde a)$
%\IF{ $\langle \mu^1,\varphi_\omega(a\oslash\tilde a) \rangle>0$}
\STATE  $a=a^{1-\omega}\otimes \tilde a^\omega$
%\ELSE
%\STATE $a=\tilde a$,\COMMENT{SK iteration}
%\ENDIF
\STATE $\tilde b=\mu_2\oslash (^t\gamma^0  a)$
\STATE $\omega=\Theta(b\oslash\tilde b)$
%\IF{ $\langle \mu^2,\varphi_\omega(b\oslash\tilde b) \rangle>0$}
\STATE  $b=b^{1-\omega}\otimes \tilde b^\omega$%\COMMENT{Overrelaxed SK iteration}
%\ELSE
%\STATE $b=\tilde b$,\COMMENT{SK iteration}
%\ENDIF
\ENDWHILE
 \RETURN $\gamma=\diag(a)\gamma^0\diag(b)$
\end{algorithmic}
\end{algorithm}

@@

Note that the local linear rate is also improved if $\theta_0$ is not the optimal choice, see \cite[chapter 5]{ciarlet1982introduction}. By remarking that for any choice of $\theta_0<2$ there is a neighborhood of $1$ on which $\varphi_{\theta_0}(\cdot)$ is nonnegative, one sees that this result applies in particular to our algorithm because eventually, for $\ell$ big enough, one has $\Theta(\gamma^\ell)=\theta_0$.
\begin{corollary}
The preceding local convergence rate applies to Algorithm \ref{thm:algo}, where $\theta_0$ is the chosen ``target'' overrelaxation parameter.
\end{corollary}
\begin{proof}
For this proof we will adopt a dual point of view, where the overrelaxed projection method boils down to the well-known SOR method~\eqref{SORdual}  and relying on the characterization $\gamma^{\ell}=e^{\alpha^\ell/\epsilon}\gamma^0e^{\beta^\ell/\epsilon}$ where $(\alpha^\ell,\beta^\ell)$ are the SOR iterates, initialized by $(\alpha^0,\beta^0)=(0,0)$. Notice that the dual problem~\eqref{DROT} has an invariant subspace due to the invariance by translation $(\alpha,\beta)\mapsto (\alpha-k,\beta+k)$, $k\in \IR$, but is strictly convex in any subspace which does not contain this line. Also, this invariance disappears in the associated primal iterates.

The SOR procedure~\eqref{SORdual}, when applied to maximize a concave quadratic form, defines a linear map $M_\omega : (\alpha^{\ell},\beta^{\ell})\mapsto (\alpha^{\ell+1},\beta^{\ell+1})$ which spectral properties are well known~\cite{ciarlet1982introduction, young2014iterative} , even in the non strictly concave case~\cite{hadjidimos1985optimization}, where convergence is replaced by \emph{semi-convergence}, i.e.\ convergence to a nonzero linear map.

>>>>

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=9cm]{images/cvgce_zone}
\caption{\label{phi_omega} Value of $\varphi_\omega(x)$. The function is positive above the black line, negative below. For any relaxation parameter $\omega$ smaller than $2$, there exists a neighborhood of $1$ on which $\varphi_\omega(\cdot)$ is positive.}
%\includegraphics[width=11cm]{phi_omega.png}
%\caption{\label{phi_omega} Plot of functions $\varphi_\omega(x)$ for different values $1\leq \omega\leq 2$ and   $x\in[10^{-2}, 10^1]$.}
\end{center}
\end{figure}

@@

\begin{proposition}\label{prop:thetachoice}
	The function
	\begin{equation}
	\label{eq:theta_k}
	\theta_k(\gamma) =\Theta\left ((A_k \gamma)\oslash \mu^k\right)
	\end{equation}
	is continuous and verifies the descent condition (\ref{eq:cond_theta_k}).
\end{proposition}
This construction, is often an excellent choice in practice, has several advantages:
\begin{enumerate}[align=right,labelsep=10pt,leftmargin=0.5cm]
\item[$\bullet$] it allows to choose arbitrarily the parameter $\theta_0$ that will be used eventually when the algorithm is close to convergence (we motivate what are good choices for $\theta_0$ in section \ref{section:local});
\item[$\bullet$] it is also an easy approach to having an adaptive method, as the approximation of $\Theta^*$ has a negligible cost (it only require to solve a one dimensional problem that depends on the smallest value of $(A_k \gamma)\oslash \mu^k$, which can be done in a few iterations of Newton's method).
%\item[$\bullet$] experiments show that this is often an excellent choice in practice.
\end{enumerate}

@@

The resulting algorithm, which is proved to be convergent by Theorem~\ref{thm:algo}, is written in pseudo-code in Algorithm \ref{SOR}. It uses the function $\Theta$ defined implicitly in \eqref{eq:Theta}, which in practice is approximated with a few iterations of Newton's method on the function $\omega \mapsto \varphi_\omega(\min u)$ which decreasing is as can be seen on Figure~\ref{phi_omega}. With the choice $\theta_0=1$, one recovers exactly the original SK algorithm.is \vspace{-0.1cm}
\begin{algorithm}
\caption{Overrelaxed SK algorithm }
\label{SOR}
\begin{algorithmic}
\REQUIRE $\mu^1\in \IR^{n_1}$, $\mu^2\in \IR^{n_2}$, $c\in \IR^{n_1\times n_2}_+$
\STATE Set $a=\mathbf{1}_{n_1}$, $b=\mathbf{1}_{n_2}$, $\gamma^0=e^{-c/\epsilon}$, $\theta_0\in[1;2)$ and $\eta>0$
\WHILE {$||a\otimes \gamma^0 b -  \mu_1||>\eta$}
\STATE $\tilde a=\mu_1\oslash (\gamma^0 b)$, 
\STATE $\omega=\Theta(a\oslash\tilde a)$
%\IF{ $\langle \mu^1,\varphi_\omega(a\oslash\tilde a) \rangle>0$}
\STATE  $a=a^{1-\omega}\otimes \tilde a^\omega$
%\ELSE
%\STATE $a=\tilde a$,\COMMENT{SK iteration}
%\ENDIF
\STATE $\tilde b=\mu_2\oslash (^t\gamma^0  a)$
\STATE $\omega=\Theta(b\oslash\tilde b)$
%\IF{ $\langle \mu^2,\varphi_\omega(b\oslash\tilde b) \rangle>0$}
\STATE  $b=b^{1-\omega}\otimes \tilde b^\omega$%\COMMENT{Overrelaxed SK iteration}
%\ELSE
%\STATE $b=\tilde b$,\COMMENT{SK iteration}
%\ENDIF
\ENDWHILE
 \RETURN $\gamma=\diag(a)\gamma^0\diag(b)$
\end{algorithmic}\vspace*{-0.1cm}
\end{algorithm}

@@

Note that the local linear rate is also improved if $\theta_0$ is not the optimal choice, see \cite[chapter 5]{ciarlet1982introduction}. By remarking that for any choice of $\theta_0<2$ there is a neighborhood of $1$ on which $\varphi_{\theta_0}(\cdot)$ is nonnegative, one sees that this result applies in particular to our algorithm because eventually, for $\ell$ big enough, one has $\Theta(\gamma^\ell)=\theta_0$.
\begin{corollary}
The preceding local convergence rate applies to Algorithm \ref{thm:algo}, where $\theta_0$ is the chosen ``target'' overrelaxation parameter.
\end{corollary}
\begin{proof}
For this proof we will adopt a dual point of view, where the overrelaxed projection method boils down to the well-known SOR method~\eqref{SORdual}  and we will rely  on the characterization $\gamma^{\ell}=e^{\alpha^\ell/\epsilon}\gamma^0e^{\beta^\ell/\epsilon}$ where $(\alpha^\ell,\beta^\ell)$ are the SOR iterates, initialized by $(\alpha^0,\beta^0)=(0,0)$. Notice that the dual problem~\eqref{DROT} has an invariant subspace due to the invariance by translation $(\alpha,\beta)\mapsto (\alpha-k,\beta+k)$, $k\in \IR$, but is strictly convex in any subspace which does not contain this line. Also, this invariance disappears in the associated primal iterates.

The SOR procedure~\eqref{SORdual}, when applied to maximize a concave quadratic form, defines a linear map $M_\omega : (\alpha^{\ell},\beta^{\ell})\mapsto (\alpha^{\ell+1},\beta^{\ell+1})$ which spectral properties are well known~\cite{ciarlet1982introduction, young2014iterative}, even in the non strictly concave case~\cite{hadjidimos1985optimization}, where convergence is replaced by \emph{semi-convergence}, i.e.\ convergence to a nonzero linear map.
In this context, when alternate maximization converges linearly at the rate $1-\eta$,  the local linear convergence rate of the SOR method is known to be $(1-\sqrt{\eta})/(1+\sqrt{\eta})$ ~\cite[Chapter 3]{chizat2017these}.
